Fast Sorting on Modern Computers

### 2005-07-16 docs/tech_report/quick.tex
fixed the quicksort predictor scripts reconsidered the special directories some changes to main (some bad, but its late and i dont care)
finsihed move, but not rename

Quicksort has been well studied, and several important improvements have been
noted. In \cite{BentleyMcIlroy93}, using a median of nine partition is discussed,
and modern discussions on quicksort refer to the problem of equal keys, which is
discussed there, as well as by Sedgewick in \cite{SedgewickBentley02}. These
improvements are not included here for several reasons. Firstly, a quicksort
which takes into account equal keys puts the equal keys at the front of the
array, then copies them into place later. These copies would represent a lot of
cache misses, and the results would overwhelm the improvements LaMarca makes in
the next phase of the algorithm. Secondly, when sorting random keys over a
large range, the number of equal keys will be vewry low. It is therefore not
worth our while to investigate this.

@@

A surprising result is that direct mapped caches often perform better than fully
associative ones. The reason for this is not clear, but it is obvious from the
graph that this is true in the case of all the quicksorts.

The branch prediction results in Figure \ref{quick branch misses} were exactly
as expected. The binary search produced a dramatic increase in the branch
misses, though not when the two-level adaptive predictor was used. A short spike
was recorded in the bimodal results, and its not clear if this is an anomaly or
not. If it is, then the two-level adaptive predictor performs worse than the
bimodal, yet again.

@@

LaMarca's results are again similar to the results presented here, but not
identical. The shape of the instruction count graph is the same in both cases,
but that of the cache misses is not. The ratio of the cache misses between base
and cache quicksort are the same in both sets of results, but the number of
cache misses in multi-quicksort is much higher here than it is in LaMarca's
findings. In addition, the shape is different. LaMarca's multi-quicksort's
cache misses plateau, while the results here show an increasing slope.

\section{Predictor results}

@@

\sixplot{base_quicksort1_0.eps}{base_quicksort_0.eps}{base_quicksort5_0.eps}{base_quicksort7_0.eps}{base_quicksort9_0.eps}{}{base quicksort without a median, with median of 3, and with pseudo median of 5, 7 and 9}

%TODO a bit to rewrite

Theres two important bits: comparison between the types of sorts, and comparisons between the different number of things tried as medians.

@@

Part 2: comparing different ways of randomly partitioning:

Ways of getting the partition:
1 - just take the rightmost element. Note that there are no sentinels, so you have to put a bounds check back in
3 - median of(0,1,2) halves
5 - median of(0, (median of(1,2,3), 4) quarters
7 - median of(median of(0,1,2), 3, median of(4,5,6)) sixths
9 - median of(median of(0,1,2), median of(3,4,5), median of(6,7,8)) eights

misprediction rates:
1 - 29.4\%
3 - 35.9\%
5 - 38.4\%
7 - 40.0\%
9 - 41.7\%

>>>>

Quicksort has been well studied, and several important improvements have been
noted. In \cite{BentleyMcIlroy93}, using a median of nine partition is discussed,
and modern discussions on quicksort refer to the problem of equal keys, which is
discussed there, as well as by Sedgewick in \cite{SedgewickBentley02}. These
improvements are not included here for several reasons. Firstly, a quicksort
which takes into account equal keys puts the equal keys at the front of the
array, then copies them into place later. These copies would represent a lot of
cache misses, and the results would overwhelm the improvements LaMarca makes in
the next phase of the algorithm. Secondly, when sorting random keys over a
large range, the number of equal keys will be vewry low. Using a median of
greater than three is discussed in section \ref{quick-predictors2}.

@@

A surprising result is that direct mapped caches often perform better than fully
associative ones. The reason for this is not clear, but it is obvious from the
graph that this is true in the case of all the quicksorts.
%TODO why is this? the only reason I've come up with is that in the direct
%ampped cache, the same area is thrashed, while keeping keys that we;re going to
%reuse is a separte bit. in the fully assoc cache, those keys get washed out by
%the keys that arent going to be reused.

The branch prediction results in Figure \ref{quick branch misses} were exactly
as expected. The binary search produced a dramatic increase in the branch
misses, when using a bimodal predictor. When using a 2-level adaptive predictor,
however, the number of misses is, in general, slightly higher, though it does
not spike due to the binary search. While this performs worse than the bimodal
predictor, it appears to show that 2-level adaptive predictors can adapt to the
erratic branches of a binary search.

@@

LaMarca's results are again similar to the results presented here, but not
identical. The shape of the instruction count graph is the same in both cases,
but that of the cache misses is not. The ratio of the cache misses between base
and cache quicksort are the same in both sets of results, but the number of
cache misses in multi-quicksort is much higher here than it is in LaMarca's
findings. In addition, the shape is different. LaMarca's multi-quicksort's
cache misses plateau, while the results here show an increasing slope.
%TODO is a pseudo-median explained
\section{Predictor results}

\subsection{Different sorts}

@@

@@

\subsection{Different medians}

\label{quick-predictors2}
\sixplot{base_quicksort1_0.eps}{base_quicksort_0.eps}{base_quicksort5_0.eps}{base_quicksort7_0.eps}{base_quicksort9_0.eps}{}{base quicksort without a median, with median of 3, and with pseudo medians of 5, 7 and 9}

%TODO a bit to rewrite
\label{quicksort entropy}

We have observed before (see Section \ref{heapsort entropy}), that the more
branches an algorithm has, the more of them will be correctly predicted. we
observe that using a better pivot reduces the number of exchanges made by
quicksort. We expect to find, however, that using a better pivot reduces the
number of correct predictions, and not the number of incorrect predictions.

Each of the \code{while} loops in \code{partition} is a branch, and each
iteration of that loop corresponds to that branch being taken, while each
exchange corresponds to that branch being not-taken. So the fewer

%TODO to whom do we owe the idea of a pseudo median? Bently/McIlroy?
%TODO when does it cut out, and why did we choose tere (benlty?)
To test this theory, we create variations of quicksort using different medians.
We tested no median (or median-of-one), median-of-three, and pseudo-medians-of
-five, -seven and -nine. A pseudo-median-of-five is achieved by choosing five
keys, evenly spaces across the array, including the first and last key. The
median of the middle three keys is determined, and the pivot is chosen as the
median of the chosen key, the first key and the last key. The median-of-seven is chosen
similarly, using seven evenly spaced keys, and finding the median of the first
three and the median of the last three, and choosing as the pivot the median of
these keys and the middle key. The median of nine determines the median of three
median-of-threes, involving the first, middle and last three keys from nine
evenly spaced keys.

We measured the number of predictions at the same points as before, aswell as
when the predictors were being calculated, expecting that a reduction in misses
in the partitioning due to using a better median would be made up for by an
increase in number of mispredictions while finding the median. The results are
shown in Figure \ref{Branch prediction performance for base quicksort without a
median, with median of 3, and with pseudo medians of 5, 7 and 9}.

Picking a better pivot reduces the number of brnahces, but doesnt reduce th
enumber of mispredictions, in fact, it increases them.
